The tree line ecotone punctuates the elevational turnover of plants, pollinators, and their interactions 

### 2021-03-12 code/ms1_beta_diversity.Rmd
a new take on ms1 analysis

### Load site data
```{r, include=FALSE, echo=FALSE}
site_data <- read_csv("../data/processed_data/site_data.csv") %>%
  dplyr::select(-c(slope.calc, slope.est, elev.min, elev.max, temp.mean)) %>% # drop these variables
  mutate(elev.class2 = factor(elev.class2, levels = c("low", "mid", "high")),
         site = factor(site)) %>% # turn elav.class into an ordered factor and site into a regular factor
  semi_join(read_csv("../data/processed_data/network.csv"), by = "site") %>% # drop sites that were not sampled
  st_as_sf(coords = c("lon", "lat")) %>%
  st_set_crs(4326) %>%
  st_transform(5684) %>%
  sfc_as_cols() %>%
  as_tibble() %>%
  dplyr::select(-geometry)
```

@@

# Patchwork plot for comparing species beta-diversity with interaction beta-diversity
png("../output/beta_stack.png", width = 6.5, height = 4.5, units = "in", res = 300)
(g1 | (g2 / g3)) + 
  plot_annotation(tag_levels = c("A", "B", "C")) + 
  #plot_layout(guides = "collect") & 
  theme_light() 
dev.off()
```

>>>>

### Load site data
```{r, include=FALSE, echo=FALSE}
site_data <- read_csv("../data/processed_data/site_data.csv") %>%
  dplyr::select(-c(slope.calc, slope.est, elev.min, elev.max, temp.mean)) %>% # drop these variables
  mutate(elev.class2 = factor(elev.class2, levels = c("low", "mid", "high")),
         site = factor(site),
         tree_line = factor(tree_line)) %>% # turn elav.class into an ordered factor and site into a regular factor
  semi_join(read_csv("../data/processed_data/network.csv"), by = "site") %>% # drop sites that were not sampled
  st_as_sf(coords = c("lon", "lat")) %>%
  st_set_crs(4326) %>%
  st_transform(5684) %>%
  sfc_as_cols() %>%
  as_tibble() %>%
  dplyr::select(-geometry)
```

@@

# Patchwork plot for comparing species beta-diversity with interaction beta-diversity
png("../output/beta_stack.png", width = 6.5, height = 4.5, units = "in", res = 300)
(g1 | (g2 / g3)) + 
  plot_annotation(tag_levels = c("A", "B", "C")) + 
  #plot_layout(guides = "collect") & 
  theme_light() 
dev.off()

pdf("../output/beta_stack.pdf", width = 6.5, height = 4.5)
(g1 | (g2 / g3)) + 
  plot_annotation(tag_levels = c("A", "B", "C")) + 
  #plot_layout(guides = "collect") & 
  theme_light() 
dev.off()
```

### 2021-03-12 ms_1/ms_1.Rmd
update
a new take on ms1 analysis

---
title: "Elevational range shifts of mountain bumble bees and their floral hosts will be punctuated by nonlinear turnover of species and interactions"

@@

Climate-driven range shifts are a fundamental evolutionary process, but the rapid pace of recent climate change has made biogeographic dynamics as much a matter of current events as of historic speculation. This provides a unique opportunity to study biogeographic processes in real time, but it also poses a grave threat to global biodiversity as species are forced to respond rapidly within the constraints of gene pools shaped by cooler climates. Climate-induced range shifts involve more than merely tracking suitable temperature envelopes; when species move, they do so in the context of biological communities and under the influence of more than just temperature, and the outcomes of climate-induced range shifts are thus contingent on potentially nonlinear processes that cannot be reduced to simple elevational or latitudinal temperature gradients. Bumble bees (Hymentopera: *Bombus* spp.) are cold-adapted species associated with mountain habitats, and they have already exhibited measurable upslope range shifts that are expected to result in habitat loss, novel competitive interactions, and disrupted pollination networks. Interpreting these shifts, however, requires an understanding of the current elevational patterns of bumble bees and the floral mutualists upon which climate change will act. In this study, we demonstrate that mountain bumble bees and flora are distributed in strongly nonlinear patterns in space and time, and that these patterns propagate from the level of species abundances to that of β-diversity, and from the level of community composition to that of biotic interactions. In particular, we show that the tree line ecotone is a critical threshold of floral resource availability, a distributional interface between high- and low-elevation bumble bees species, and a zone of rapid acceleration of species and interaction β-diversity. The implications of these findings extend beyond the particular case of bumble bees to demonstrate that linear elevational temperature gradients are ecologically punctuated, and the outcomes of climate-induced range shifts with depend on how species and species-interactions change as they cross --- or fail to cross --- thresholds like the tree line ecotone.

@@

The rate of recent climate change is forcing species to adapt ecologically within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher latitudes and higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Species range shifts, however, are not merely abstract, Cartesian translations along axes of latitude, elevation, and temperature; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Moreover, some of these factors, unlike temperature, cannot be assumed to vary linearly with latitude or elevation but instead exhibit marked nonlinearities and thresholds. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a continuous elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir]. Thus, species range shifts --- and, perhaps more importantly, the ecological consequences thereof --- are contingent upon existing patterns of species distributions and biotic interactions, and community ecology is as much prior as posterior to the ecology of global change.

Bumble bees (Hymenoptera: *Bombus* spp.) are both classic model taxa in community ecology and an emergent conservation priority due to recent declines [@Goulson2008-oe] and manifest vulnerability to climate change [@Soroye2020-zb]. The genus *Bombus* is believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary (~34 mya), and its subsequent spread and diversification along mountain corridors and into lowland habitats was likely driven by alternating range expansion during cooling periods and retreat to higher elevations during warming periods [@Hines2008-er; @Stewart2010-jn; @Martinet2018-rb]. Today, bumble bees remain characteristically cold-adapted species [@Heinrich1994-cv], with peak abundance and diversity associated with mountain ranges and northern latitudes, where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Climate-induced range shifts are expected to be especially pronounced in the mountain habitats that host the core of bumble bee biodiversity, together with their floral mutualists, and upward movement of bumble bee species has already been documented [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. Populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer geometric habitat shrinkage and increasing genetic isolation as they move toward mountain peaks; the upward advance of lowland species may introduce novel competitive pressures and disrupt the pollination networks in which bumble bees participate [@Brosi2013-mu; @Ishii2013-sm]; and concurrent elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their historic floral hosts.

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective, but they also provide a model system in which to study the general phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. Anticipating, interpreting, and potentially mitigating the effects of ranges shifts, however, hinge upon an understanding the current elevational and temporal patterns of the bumble bee and floral communities that will be acted upon by climate change. In particular, it is important to detect nonlinearities in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id].

In the present study, we approach this topic at three levels of biological organization. First, we ask how mountain bumble bee species are distributed in elevation and intra-annual time with respect to one another and with respect to their floral hosts. Then, we move from the level of individual species to that of community composition and explore the structure of species turnover (i.e. β-diversity) of bumble bees and flora in response to elevation. Finally, we move to the level of species interactions, investigating the turnover of bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques capable of capturing nonlinear responses with the goal of identifying where, when, and for which species the impacts of climate-induced range shifts should be expected to be most severe

@@

The study was conducted 2010-2012 in Berchtesgaden National Park (47.55°N, 12.92°E), located in the Northern Limestone Alps of southeast Germany. The landscape is composed of mountain pastures mainly surrounded by coniferous forests. We selected 25 study sites (60 x 60 m) on mountain pastures at elevations ranging from 641-2032 m above sea level. Fourteen of these pastures are extensively grazed by cattle or sheep, three are mowed for hay production, and eight have been abandoned throughout the last century and are no longer subject to any human management. The sites sampled in this study are the same as those used in a series of previous studies [@Hoiss2012-sn; @Hoiss2012-kr; @Hoiss2015-gv], but the data reported here are otherwise independent.

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected on dry days when the air temperature was at least 6°C. During each visit, bumble bees and their activity were recorded during a 50-minute transect walk. Bumblebees observed on or a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory after [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers. In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.1 m^2^. Species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

While per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) varied markedly,  with 153 in 2010, 212 in 2011, and 539 in 2012. The increased total sampling effort in 2012 was driven in part by the overall scarcity of bumble bees that year. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient. See Appendix X for details of sampling patterns.

@@

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date (excluding males). To represent floral abundance from the perspective of each bumble bee species in our network, we first scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of visited genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection on the estimation of floral resource abundance, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp]. We also opted to treat the cryptic *B. terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be distinguished reliably in the field. Similarly, we pooled representatives of the parasitic *Psithyrus* subgenus --- which included *B. barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into a single species group *B. psithryus*.

We analyzed bumble bee abundance and floral abundance using separate hierarchical generalized additive models (HGAM) [@Wood2011-pg; @Pedersen2019-kf], with abundance modeled as a response to the interaction (tensor product) of elevation and day-of-year. Bumble bee species was included in each model as both a smoothing factor and a random intercept term, resulting in centered 2-dimensional abundance smooths for each bumble bee species. We also included year and site as random intercept effects to account for annual differences in abundance patterns and repeated measurements within sites.

To complement GAM-based analysis of abundance patterns through elevation and time, we also analyzed the site occupancy pattern of each bumble bee species using cluster and ordination analyses. We first calculated the mean abundance across dates of each bumble bee species for each site in our study system. Then, we applied a Hellinger transformation to the matrix of mean abundances and calculated the Horn-Morisita distance metric for each pair of species. The resulting distance matrix was then used to generate a principal coordinates analysis (PCoA) and a cluster dendrogram, each depicting the (dis)similarity between species pairs in their distribution across sites. The primary purpose of this analysis was to organize the presentation of the GAM results by grouping species with similar site affinities, but because our sites were arranged along an elevation gradient, our cluster and ordination analyses also highlight differences between species in elevational distribution.

For our analyses of species and interaction β-diversity, we opted to focus on elevational patterns by pooling samples within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated the species and interaction β-diversity between all pairs of sites and partitioned total interaction β-diversity into its components of species turnover (further partitioned into bumble bee and plant turnover, respectively) and interaction rewiring [@Novotny2009-my; @Poisot2012-fk].

As an initial analysis, we plotted the relationship between site-wise elevation difference and each metric of β-diversity and verified the significance of the relationship using logistic matrix regression [@Goslee2009-ln], with the difference in number of sampling dates between sites as a covariate to control for potential confounding effects of sampling frequency.

@@

All analyses were conducted in `R` `r citep(citation())`. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa], and the models we fit correspond to the “type I” model form described in @Pedersen2019-kf. PCoA and species β-diversity calculation were performed with the package `vegan` `r citep(citation("vegan"))`. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. Matrix regression was performed with the package `ecodist` [@Goslee2007-aa]. GDM analysis of β-diversity was performed with the package `gdm` `r citep(citation("gdm"))`. Data handling and visualization was performed with the `tidyverse` suite of R packages `r citep(citation("gdm"))`. Annotated R code is available in the Supplementary Material.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris-lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites **(Figure 1)**. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily above 1500 m, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across mid-elevation sites. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses.

```{r fig1, echo = FALSE,  fig.cap = paste("Mean abundances and elevational distributions of bumble bee species. Sampled points along the elevation gradient are depicted as circles, the diameter of which is scaled to the mean abundance (pooled across study years) of the given bumble bee species at that elevation. Abundances of zero are not plotted. Lines span between the highest and lowest occurrence of each species, highlighting its observed elevation range.")}
knitr::include_graphics("../output/bb_range.png")
```

@@

## Abundance over elevation and time

@@

```{r fig2, echo = FALSE,  fig.cap = paste("Abundance of bumble bee species and their floral resources through elevation and time. For each bumble bee species, surface plots are arranged in two columns labeled BB and FL. Surface plots in the BB columns show the abundance of the bumble bee species, while plots in the FL columns show the abundance of floral resources for the corresponding bumble bee species. The color ramp of each smooth has been scaled to equal color range, thereby obscuring the absolute magnitude of effect. To avoid cluttering the figure, we have opted not to include a unique color ramp key for each plot, since the focus of our inference is on relative rather than absolute patterns. Surface plots with original color ramp scales are available in Appendix X.")}
knitr::include_graphics("../output/gam_BB_FL4.png")
```

@@

```{r fig3, echo = FALSE,  fig.cap = paste("Species-level (A) and interaction-level (B, C) plotted against elevational difference. Each point represents the β-diversity (or β-diversity partition) between a pair of sites, and lines represent the overall relationship between β-diversity and elevation difference using binomial regression smooths. Standard errors are not plotted because they would be misleading due to the non-independence inherent to distance matrix regression, but all regressions were significant (p < 0.01). Interaction β-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned β-diversity, ST = β-diversity due to species turnover, OS =  β-diversity due to interaction rewiring, ST.h =  β-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = β-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  β-diversity due to joint species turnover in higher and lower trophic levels.")}
knitr::include_graphics("../output/beta_stack.png")
```

@@

```{r fig4, echo = FALSE,  fig.cap = paste("GDM splines of species-level and interaction-level β-diversity in response to elevational, controlling for geographic proximity and number of sampling dates. The maximum height of each spline represents a given variable’s partial effect on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the corresponding gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Within-site comparisons were omitted from GDM analysis to avoid pseudoreplication. Shaded bands depict the uncertainty of each partial effect curve (+/- one standard deviation) based on bootstrapping. Note the differences in scale between the y-axes of each variable.")}
knitr::include_graphics("../output/gdm_stack.png")
```

>>>>

---
title: "The tree line ecotone punctuates the elevational turnover of plants, pollinators, and their interactions "

@@

Climate-induced range shifts involve more than merely tracking suitable temperature envelopes; when species move, they do so in the context of biological communities and under the influence of more than just temperature, and the outcomes of climate-induced range shifts are thus contingent on processes that cannot be reduced to linear elevational or latitudinal temperature gradients. Bumble bees (Hymentopera: *Bombus* spp.) are cold-adapted species associated with mountain habitats, and they have already exhibited measurable upslope range shifts that are expected to result in habitat loss, novel competitive interactions, and disrupted pollination networks. Interpreting these shifts, however, requires an understanding of the current elevational patterns of bumble bees and the floral mutualists upon which climate change will act. In this study, we analyze bumble bee, flowering plant, and interaction records sampled during 904 standardized transect walks conducted over three consecutive years along an elevational gradient in the German Alps ranging from 640 to 2030 meters above sea level.

We demonstrate that mountain bumble bees and flora exhibit distinct peaks of abundance, are distributed in strongly nonlinear patterns in space and time, and that these patterns propagate from the level of species abundances to that of β-diversity, and from the level of community composition to that of biotic interactions. In particular, we show that the tree line ecotone is a critical threshold of floral resource availability, a distributional interface between high- and low-elevation bumble bees species, and a zone of rapid acceleration of species and interaction β-diversity. The implications of these findings extend beyond the particular case of bumble bees to demonstrate that linear elevational temperature gradients are ecologically punctuated, and the outcomes of climate-induced range shifts with depend on how species and species-interactions change as they cross --- or fail to cross --- thresholds like the tree line ecotone.

@@

The rate of recent climate change is forcing species to adapt ecologically within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher latitudes and higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Species range shifts, however, are not merely abstract translations along axes of latitude and elevation; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Thus, when species move along continuous, linear climate gradients, they can experience punctuated and nonlinear ecological transitions. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a continuous elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir]. Thus, species range shifts and the consequences thereof are contingent upon existing patterns of species distributions and biotic interactions.

Bumble bees (Hymenoptera: *Bombus* spp.) are well-studied model organisms in the field of community ecology, but they have more recently become conservation priorities due to well-documented population declines and particular vulnerability to climate change [@Goulson2008-oe; @Soroye2020-zb]. The genus *Bombus* is believed to have arisen in the mountains of Asia during the cooling climate of the Eocene-Oligocene boundary (~34 mya) [@Hines2008-er], and bumble bees remain characteristically cold-adapted species today [@Heinrich1994-cv], with peak abundance and diversity in mountain ranges and northern latitudes where they are often the dominant guild of flower-visiting insects and the principal pollinators of entomophilous flora [@Williams1998-xi; @Goulson2010-lo]. Climate-induced range shifts are expected to be especially pronounced in the mountain habitats that host the core of bumble bee biodiversity, together with their floral mutualists, and upward movement of bumble bee species has already been documented [@Ploquin2013-da; @Kerr2015-dg; @Pyke2016-ll; @Fourcade2019-ct; @Soroye2020-zb; @Marshall2020-pc]. Populations of bumble bee species already restricted to the highest elevation zones can be expected to suffer geometric habitat shrinkage and increasing genetic isolation as they move toward mountain peaks; the upward advance of lowland species may introduce novel competitive pressures and disrupt the pollination networks in which bumble bees participate [@Brosi2013-mu; @Ishii2013-sm]; and concurrent elevational [@Lenoir2008-yn] and/or phenological [@Pyke2016-ll] shifts in the floral community could generate mismatches between bumble bee species and their historic floral hosts.

<!-- I either need to beef up the paragraph below or fold it into another one. But I need to get to the point of our study. Which means I need to know what the point is. -->

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic phenomenon of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. As bumble bee species move up mountain slopes, they move in relation to plant species and other bumble bee species

Anticipating, interpreting, and potentially mitigating the effects of ranges shifts, however, hinge upon an understanding the current elevational and temporal patterns of the bumble bee and floral communities that will be acted upon by climate change. In particular, it is important to detect nonlinearities in species-, community-, and interaction-level responses to elevational climate gradients, such as the floristic threshold effect that has been documented at the tree line ecotone [@Descombes2017-id].

⎄
In the present study, we approach this topic at three levels of biological organization. First, we ask how mountain bumble bee species are distributed in elevation and intra-annual time with respect to one another and with respect to their floral hosts. Then, we move from the level of individual species to that of community composition and explore the structure of species turnover (i.e. β-diversity) of bumble bees and flora in response to elevation. Finally, we proceed to the level of species interactions, investigating the turnover of bee-flower interaction partners through elevation. In each of these analyses, we employ flexible modeling techniques capable of capturing nonlinear responses with the goal of identifying where, when, and for which species the impacts of climate-induced range shifts should be expected to be most severe.

@@

The study was conducted 2010-2012 in Berchtesgaden National Park, located in the Northern Limestone Alps of southeast Germany (47.55°N, 12.92°E). The landscape is composed of mountain pastures mainly surrounded by coniferous forests. We selected 25 study sites (60 x 60 m) on mountain pastures at elevations ranging from 641-2032 meters above sea level (m.s.l.) **(Figure 1)**. Fourteen of these pastures are extensively grazed by cattle or sheep, 3 are mowed for hay production, and 8 have been abandoned throughout the last century and are no longer subject to any human management. Study sites were classified as falling either above or below the tree line based on an examination of Google Earth imagery and corroborated by the authors with field experience at our study sites (K. Kallnik, A. F. Maihoff, A. Claßen). Among our sites, the tree line generally fell at an elevation of around 1500 m.s.l., consistent with the previous descriptions of the tree line in the Berchtesgander Alps [@Kostler1970-ya; @Mayer1970-zn] **(Figure 2)**. Eight sites fell above the tree line and 17 fell below it **(Figures 1, 2)**.

```{r fig1, echo = FALSE,  fig.cap = paste("Study sites in the Berchtesgadener Alps plotted over Google Earth imagery. Sites above the tree line are depecited in red and sites below the tree line in yellow.")}
knitr::include_graphics("../sites2_2021-03-11_17-40-25.png")
```

Sampling in each year of the study consisted of repeated visits to each site at approximately weekly intervals. Samples were only collected during periods without rain when the air temperature was at least 6°C. During each visit, bumble bees and their activity were recorded during a 50-minute transect walk. Bumblebees observed on or a given flower were counted as floral visitors. Bumble bee queens were identified to species level in the field, while workers and males were stored in individually labeled tubes in the freezer for later identification in the laboratory after [@Amiet1996-jz]. Floral visitation by male bumble bees was recorded during visitation sampling, but we chose to analyze only visitation by queens and workers; males are often found resting on flowers, so their "visits" cannot necessarily be interpreted in the same way as those of queens and workers. In conjunction with visitation observations, we estimated the flower cover of each herbaceous or shrubby plant species within each 60 x 60 m study plot to the nearest 0.01 m^2^. Species identification followed @Lauber2007-ha and @Oberndorfer2001-ro.

Although per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) varied: 153 in 2010, 212 in 2011, and 539 in 2012. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient. See Appendix X for details of sampling patterns.

@@

### Data processing

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date. To represent floral abundance from the perspective of each bumble bee species in our network, we first scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of visited genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection on the estimation of floral resource abundance, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp]. We also opted to treat the cryptic *B. terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished reliably. We also pooled representatives of the parasitic *Psithyrus* subgenus --- which included *B. barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into a single species group *B. psithryus*.

### Distribution of bumble bees and floral resources in elevation and time

We analyzed bumble bee abundance and floral abundance using separate hierarchical generalized additive models (HGAM) [@Wood2011-pg; @Pedersen2019-kf], with abundance modeled as a response to the interaction (tensor product) of elevation and day-of-year. Bumble bee species was included in each model as both a smoothing factor and a random intercept term, resulting in centered 2-dimensional abundance smooths for each bumble bee species. We also included year and site as random intercept effects to account for annual differences in mean abundance and repeated measurements within sites.

### Species and interaction β-diversity

For our analyses of species and interaction β-diversity, we opted to focus on elevational patterns by pooling samples within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated the species and interaction β-diversity between all pairs of sites and partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (identical species interacting in different ways) [@Novotny2009-my; @Poisot2012-fk]. As an initial analysis, we plotted the relationship between site-wise elevation difference and each metric of β-diversity and verified the significance of the relationship using logistic matrix regression [@Goslee2009-ln], with the difference in number of sampling dates between sites as a covariate to control for potential confounding effects of sampling frequency.

@@

### Software

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa], and the models we fit correspond to the “type I” model form described in @Pedersen2019-kf. PCoA and species β-diversity calculation were performed with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. Matrix regression was performed with the package `ecodist` [@Goslee2007-aa]. GDM analysis of β-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. Annotated `R` code is available in the Supplementary Material.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. Five species --- *B. pascuorum*, *B. pratorum*, *B. soroeensis*, *B. terrestris/lucorum*, and *B. wurflenii* --- accounted for the bulk of overall bumble bee abundance and were present over the entire elevation range of our sites **(Figure 1)**. *B. hortorum*, *B. jonellus*, and *B. psithyrus* (species-group) were widespread but at lower abundance. *B. monticola* occurred at moderate abundance but was rare or absent at sites below 1000 m.s.l. The remaining species were relatively rare in our study system. *B. mendax*, *B. mucidus*, and *B. pyrenaeus* occurred primarily (peak abundance > 1) above the 1500 m.s.l., consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. *B. gerstaeckeri*, *B. hypnorum*, *B. lapidarius* were sparsely distributed across low- and mid-elevation sites. *B. humilis* was recorded only three times and only in one year, so we omitted it from all analyses.

```{r fig2, echo = FALSE,  fig.cap = paste("Peak abundances and elevational distributions of bumble bee species. Sampled sites along the elevation gradient are depicted as circles, the diameter of which is scaled to the peak abundance (pooled across study years) of the given bumble bee species. Abundances of zero are not plotted. Lines span between the highest and lowest occurrence of each species, highlighting its observed elevation range. Species are plotted in order of increasing median elevation.")}
knitr::include_graphics("../output/bb_range_ms1.png")
```

@@

## Distribution of bumble bees and floral resources in elevation and time

@@

```{r fig3, echo = FALSE,  fig.cap = paste("Abundance of bumble bee species and their floral resources through elevation and time. For each bumble bee species, surface plots are arranged in two columns labeled BB and FL. Surface plots in the BB columns show the abundance of the bumble bee species, while plots in the FL columns show the abundance of floral resources for the corresponding bumble bee species. The color ramp of each smooth has been scaled to equal color range, thereby obscuring the absolute magnitude of effect. To avoid cluttering the figure, we have opted not to include a unique color ramp key for each plot, since the focus of our inference is on relative rather than absolute patterns. Surface plots with original color ramp scales are available in Appendix X.")}
knitr::include_graphics("../output/gam_BB_FL4.png")
```

@@

```{r fig4, echo = FALSE,  fig.cap = paste("Species-level (A) and interaction-level (B, C) plotted against elevational difference. Each point represents the β-diversity (or β-diversity partition) between a pair of sites, and lines represent the overall relationship between β-diversity and elevation difference using binomial regression smooths. Standard errors are not plotted because they would be misleading due to the non-independence inherent to distance matrix regression, but all regressions were significant (p < 0.01). Interaction β-diversity (B, C) is partitioned using Poisot et al.’s (2012) notation: WN = unpartitioned β-diversity, ST = β-diversity due to species turnover, OS =  β-diversity due to interaction rewiring, ST.h =  β-diversity due to species turnover in the higher trophic level (bumble bees), ST.l = β-diversity due to species turnover in the lower trophic level (plants), and ST.lh =  β-diversity due to joint species turnover in higher and lower trophic levels.")}
knitr::include_graphics("../output/beta_stack.png")
```

@@

```{r fig5, echo = FALSE,  fig.cap = paste("GDM splines of species-level and interaction-level β-diversity in response to elevational, controlling for geographic proximity and number of sampling dates. The maximum height of each spline represents a given variable’s partial effect on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the corresponding gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Dotted vertical lines in the elevation panels indicates the location of the tree line. Within-site comparisons were omitted from GDM analysis to avoid pseudoreplication. Shaded bands depict the uncertainty of each partial effect curve (+/- one standard deviation) based on bootstrapping. Note the differences in scale between the y-axes of each variable.")}
knitr::include_graphics("../output/gdm_stack.png")
```
